### 2023-04-20 paper.Rmd
editing

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means anticlustering is used to partition $N$ elements each having $P$ attributes into $K$ groups $(C_1, \ldots, C_K)$. The partitioning is subject to the requirement that each element is assigned to exactly one group. In anticlustering applications, an additional constraint is usually imposed on the group sizes, with the most common restriction being that all groups have equal size [@papenberg2020].

@@

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all $n_k$ observations belonging to the $k$'th cluster $C_k$ ($k = 1, \ldots, K$):

@@

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variance between groups instead of only means. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, we compute new variables for each original variable and for each distribution moment that should be considered in the anticlustering process.

To formalize the computation of the k-plus criterion for an arbitrary number of distribution moments, some new notation has to be provided. First, we define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be equalized between $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{.j}$, a new *t-moment variable* $\hat{x}_{t.j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ again has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional $t$-moment variables, appending them as new columns to the original data matrix, and computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing $\mathit{SSE_{kplusGeneralized}}$. Because the computation of the new $t$-moment variables involves exponentiation, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of the different moments when computing $\mathit{SSE_{kplusGeneralized}}$. To ensure a comparable weight of all criteria, all attributes---i.e., the original data as well as the new $t$-moment variables---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization. This approach will henceforth be referred to as the unweighted standardization approach for k-plus anticlustering.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented variables---of the input data $\mathbf{X'}_{N \times \mathit{TP}}$ before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized matrix $\mathbf{X'}_{N \times \mathit{TP}}$ was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, repectively. The weighted approaches did not use a standardization of the input data.

@@

The function call takes about 10 seconds on a contemporary personal computer. The function `kplus_anticlustering()` has three boolean arguments to specify whether variance, skewness and kurtosis should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness and kurtosis have to be specifically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features are standardized before the optimization starts; in particular, this option enforces that all k-plus criteria receive the same weight during the optimization process. This behaviour can be adjusted using the boolean argument `standardize`.

>>>>

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means anticlustering is used to partition $N$ elements each having $P$ attributes into $K$ groups $(C_1, \ldots, C_K)$. The partitioning is subject to the requirement that each element is assigned to exactly one group. In anticlustering applications, an additional constraint is usually imposed on the group sizes $n_k$ ($k = 1, \ldots K$), with the most common restriction being that all groups have equal size, i.e. $n_k = \frac{N}{K}$ [@papenberg2020].

@@

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all observations belonging to the $k$'th cluster $C_k$ ($k = 1, \ldots, K$):

@@

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variances between groups instead of only means. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, we compute new variables for each original variable and for each distribution moment that should be considered in the anticlustering process. Analogously to the computation of, we could then define a $\mathit{SSE_{Skew}}$ for $t = 3$ or a $\mathit{SSE_{Kurtosis}}$ for $t = 4$. However, to formalize the computation of the k-plus criterion for an arbitrary number of distribution moments, some new notation has to be provided.

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be equalized between $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{.j}$, a new *t-moment variable* $\hat{x}_{t.j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional $t$-moment variables, appending them as new columns to the original data matrix, and computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing $\mathit{SSE_{kplusGeneralized}}$. Because the computation of the new $t$-moment variables involves exponentiation, their range of values is expected to be very different from the original variables. This imbalance can strongly and unpredictably influence the relative weight of the different moments when computing $\mathit{SSE_{kplusGeneralized}}$ (e.g., see Figure 1). To ensure a comparable weight of all criteria, all attributes---i.e., the original data as well as the new $t$-moment variables---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization. This approach will henceforth be referred to as the unweighted standardization approach for k-plus anticlustering.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented variables---of the input data $\mathbf{X'}_{N \times \mathit{TP}}$ before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized matrix $\mathbf{X'}_{N \times \mathit{TP}}$ was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

The function call takes about 10 seconds on a contemporary personal computer. The function `kplus_anticlustering()` has three boolean arguments to specify whether variance, skewness and kurtosis should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness and kurtosis have to be specifically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features---the original and the augmented variables---are standardized before the optimization starts. Thus, by default, the unweighted standardization approach is used. Whether standardization is used can be adjusted using the boolean argument `standardize`.
